Reddit /r/Depression Authors who Pen Suicidal Ideation Leave Reddit at Higher Rates

### 2016-10-27 Paper_Rough_Draft.Rmd
edited rough draft. added supplement.

@@

The goal for our classification algorithm is to determine whether comments and posts contain expressions of suicidal ideation. From these classifications, we produce a set of authors that express SI in June /r/Depression.

The CDC defines suicidal ideation, or thoughts about suicide, as "thoughts of engaging in suicide-related behavior" [cite] and "Thinking about, considering, or planning suicide." [cite]. We broaden this definition for our analysis to the following:

      Suicidal Ideation: Thinking about, considering, or planning suicide; desire to die; desire to stop living; or engagement in highly dangerous activities or self harm with the implicit goal of ending life. Suicidal ideation does not include hating or frustration with one's own life, extreme anguish or psychological distress, or engagement in highly dangerous activities or self harm where there is no implicit goal of ending one's life.

It is important to note that we expected SI expression to be fairly rare among /r/Depression posts. Thus, in order to get a good estimate of the difference in drop-out rates between true SI-positive and true SI-negative authors, we wanted our SI-positive sample to have a minimal proportion of true SI-negative authors. However, we were not worried about true SI-positive authors tainting our SI-negative sample since we expected SI-positive users to be so rare. In other words, we wanted a highly specific algorithm at the sacrifice of sensitivity.

We read through a sample of 750 randomly selected comments and posts from June /r/Depression, and we classify SI according to the definition above. From each of the SI-positive comments, we extract the smallest phrase(s) that contained evidence of SI. Examples of these phrases include:

* "I think about suicide"

* "I just don't want to live""

* "I have struggled with depression and suicidal thoughts"

* "my suicide plan"

* "I still feel like blowing my brains out"

We then categoize collected phrases into twelve linguistically distinct modes of expressing SI. We discarded phrases that we felt were too linguistically complex for any classification algorithm to handle. The twelve modes are:

* [I am] [intending to] [suicide verb]

* [I am] [thinking about] [suicide noun]

* I [intend to] [suicide verb]

* I [suicide verb]

* [I am] [suicide adjective]

* Why not [suicide verb]

* I [think about] [suicide noun]

* me [suicide verb]

* me [suicide noun]

* I [oppose] [live noun]

* I [am opposed to] [living]

The phrases or words in brackets represent a broader class of phrases that could be inserted in it's place. For example, [suicide verb] could be replaced with any of the following phrases to still consitute evidence of SI expression:

@@

For each of these modes of SI expression, we generate all possible combinations of phrases. After all phrases for all 12 modes of SI expression have been generated, we remove all punctuation and spacing from both the post text and the phrase text. We combine all phrases into a regular expression using the or "|" operator, and then use the function "grepl" to test whether each post contains any phrase.

@@

### June /r/depression

For the month of June, /r/depression contained 22,041 posts, representing contributions from 7,572 authors. We identified 547 SI-positive posts from 492 unique users.

In a sample of 100 SI-positive posts, [n] did not have true SI evidence. In a sample of 100 SI-negative posts, [n] did have SI evidence.

@@

```{r , echo=FALSE, message=FALSE, warning=FALSE, fig.width=10}
user.data = 
      author.summary %>%
      group_by(author) %>%
      filter(newest == max(newest)) %>%
      ungroup %>%
      group_by(si.author)%>%
      transmute(author, rank_newest = rank(newest)) %>%
      ungroup %>%
      select(-si.author) %>%
      merge(user.data, by = "author", all.y=T)
user.data$point.label = 
      ifelse(user.data$class, "SI",
      ifelse(user.data$r.depression, "/r/depression","other")) %>%
      ordered(labels = c("other", "/r/depression", "SI"))
user.data$si.group = 
      ifelse(user.data$si.author, "SI-Positive in June /r/depression",
             "SI-Negative in June /r/depression")
### Make plot
ggplot(data = subset(user.data, date > jan1)) +
geom_point(aes(y=rank_newest,x=date, 
               color=point.label, alpha=point.label, 
               size=point.label)) + 
ggtitle("Distribution of Post Dates across Authors\nbetween Jan. and Oct.") +
ylab("Author Rank by Newest Post") + xlab("Date") +
geom_vline(aes(xintercept = as.numeric(july1)), alpha=0.3) + 
geom_vline(aes(xintercept = as.numeric(june1)), alpha=0.3) + 
scale_colour_manual(values = c("grey","darkblue", "red")) + 
scale_size_manual(values = c(0.1,0.1,0.3)) + 
scale_alpha_manual(values = c(0.1,0.1,1)) + 
theme_bw() + facet_wrap("si.group")
```

After scraping the post history for each of the SI-positive users, we find no post or comment data for 1 author, and we find 16 authors whose full history was not available back to April 2016. We are left with 475 authors.

The post histories for our 475 authors are depicted above. Each y value represents a unique author id, sorted by the time of their most recent post. Each x value represents the date of a post. Hence, we can read each horizontal line on this plot as the post history of one author. Red points represent SI-positive posts. Blue points represent posts in the /r/depression subreddit. Grey posts represent posts elsewhere on Reddit. There a number of important features in this plot. First, we note that those authors who posted more recently tended to post more often, which is expected. Second we see that many authors made multiple SI-positive posts. Third, we see that nearly 170 of our 475 authors had their last post in June, which was the period that we used to sample from. Fourth, of these 170 authors, many did not post frequently before dropping-off Reddit. However, a hand-full of them had posted frequently, and their drop-put is therefore suspicious.

@@

```{r, echo=FALSE, warning=FALSE}
### Load Data
      author.summary = 
            user.data %>% 
            group_by(author) %>%
            summarise(n = n()-1,
                      eps.bar = mean(time.last, na.rm=T),
                      t = as.numeric(october10 - max(date)),
                      si.author = si.author[1])
            filter(author.summary, !t < 150)
      n = author.summary$n
      eps = author.summary$eps.bar
      eps[is.na(eps)] = 0 
      t = author.summary$t
      N = nrow(author.summary)
      si = author.summary$si.author
### Simulate Mu given Data
      alpha = 1; beta = 1
      mu = replicate(n = 10000, rgamma(N, shape = alpha + n, rate = beta + n*eps))
### Now get overall credible interval for sum(Q)
      prior.p = 10^-10
      make.p = function(mu,p,t) p / (exp(-mu*t)*(1-p) + p)
      post.si = colSums(apply(mu[si,], 2, make.p, p=prior.p, t=t[si]))
      post.nosi = colSums(apply(mu[!si,], 2, make.p, p=prior.p, t=t[!si]))/sum(!si)*sum(si)
      post.excess = post.si - post.nosi
      hist(post.excess, main = "Posterior Distribution for\nExpected Reddit Drop-Outs per 100 SI-Positive Authors\n(Prior = 10e-4)",
           xlab="", ylab="")
      ci = quantile(post.excess, probs = c(0.025, 0.975))
```

@@

We found that our conclusions were highly sensitive to our specified prior probability of drop-outs. The plot below shows posterior credible intervals for the expected number of excess dropouts among authors who penned SI-positive posts in June /r/depression.

@@

```{r, echo=FALSE, warning=FALSE}
### Calculate P(Q) from Mu and data
      prior.p = 10^-seq(0.5,15.5, by=0.5)
      post_p = sapply(prior.p, FUN = function(x){
            post.si = colSums(apply(mu[si,], 2, make.p, p=x, t=t[si]))
            post.nosi = colSums(apply(mu[!si,], 2, make.p, p=x, t=t[!si]))/sum(!si)*sum(si)
            post.excess = post.si - post.nosi
            return(quantile(post.excess, probs = c(0.025, 0.5, 0.975)))
      })
      colnames(post_p) = paste0("p10e-", seq(0.5,15.5, by=0.5))
      #post_p = cbind(Percentile = rownames(post_p), as.data.frame(post_p))
      #post_p = melt(post_p, id.vars = "Percentile", 
      #              variable.name = "Prior P",
      #              value.name = "Estimate")
      ggplot() +
            geom_errorbar(aes(ymax = post_p[3,], ymin = post_p[1,], x=prior.p)) +
            geom_point(aes(y= post_p[2,], x=prior.p)) + 
            scale_x_log10() + 
            xlab("Prior Probability of Quitting Reddit") + 
            ylab("Excess Droup-Outs Associated with SI") +
            ggtitle("Posterior Credible Intervals for Total SI-Associated Drop-Outs") 
```

>>>>

Collaborators: Linda Gai

Peer Editors: Guanghao Qi, Matt Cole

Supplemental Materials: [Supplement](Supplement.html).

@@

##### SI Definition

The goal for our classification algorithm is to determine whether comments and posts contain expressions of suicidal ideation (SI). From these classifications, we produce a set of authors that express SI in June /r/Depression.

The CDC defines suicidal ideation, or thoughts about suicide, as "thoughts of engaging in suicide-related behavior" [cite] and "Thinking about, considering, or planning suicide." [cite]. For our analysis, we change this definition to the following:

We define suicidal ideation (SI) as thinking about, considering, or planning suicide; desire to die; desire to stop living; or engagement in highly dangerous activities or self harm with the implicit goal of ending life. It may include admissions of having suicidal ideation in the past or the potential for suicidal ideation in the future. Evidence of suicidal ideation does not follow from hating or frustration with one's own life, extreme anguish or psychological distress, or engagement in highly dangerous activities or self harm where there is no implicit goal of ending one's life. However, explicit statements about wanting to die or engage in suicide are taken literally.

We change our definition of suicidal ideation to make it clearer what kinds of Reddit comments may constitute SI. For example, the phrase "I don't want to live." does not fall under the CDC SI definition since there is no explicit thoughts of lethal self-harm. Rather, this statement expresses SI through a double-negative. While this definition does not fit a traditional definition of SI, we include these kinds of comments because Reddit moderators may be equally worried about authors who express these comments.

##### The Algorithm

A number of methods have been implemented to detect suicidal ideation from user-generated text on social media [CITE]. However, for simplicity, we do not use them here.

We expected SI expression to be fairly rare among /r/Depression posts. Thus, in order to get a good estimate of the drop-out risk attributable to SI, we wanted our SI-positive sample to have a minimal proportion of true SI-negative authors. However, we were not worried about true SI-positive authors tainting our SI-negative sample since we expected SI-positive author to be so rare, and hence not have much influence on our estimate of the SI-negative drop-out risk. In other words, we wanted a highly specific algorithm at the sacrifice of sensitivity.

With an intent to achieve this goal, we design an algorithm that searches post text for at least one phrase from an explicit list of SI-expressive phrases that constitute strong evidence of suicidal ideation [Supplement Reference].

These phrases were derived from a random sample of 750 June 2016 /r/Depression comments. Human classifiers read through each comment, labeled whether it contained SI expression, and recorded the smallest phrase(s) from that comment that the classifier felt constituted evidence of SI. We then organized the phrases into twelve categories, and noted in brackets where words could be exchangable. For example, [suicide verb] could be replaced with any of the following phrases to still consitute evidence of SI:

@@

@@

### June /r/Depression

For the month of June, /r/depression contained 22,041 posts, representing contributions from 7,572 authors. We identified 547 SI-positive posts from 492 unique authors. In a simple random sample of 100 SI-positive posts, 12 did not have true SI evidence. In a simple random sample of 100 SI-negative posts, 4 did have true SI evidence.

@@

```{r , echo=FALSE, message=FALSE, warning=FALSE, fig.width=10}
user.data = 
      author.summary %>%
      group_by(author) %>%
      filter(newest == max(newest)) %>%
      ungroup %>%
      group_by(si.author)%>%
      transmute(author, rank_newest = rank(newest)) %>%
      ungroup %>%
      select(-si.author) %>%
      merge(user.data, by = "author", all.y=T)
user.data$point.label = 
      ifelse(user.data$class, "SI",
      ifelse(user.data$r.depression, "/r/depression","other")) %>%
      ordered(labels = c("other", "/r/depression", "SI"))
user.data$si.group = 
      ifelse(user.data$si.author, "SI-Positive in June /r/depression",
             "SI-Negative in June /r/depression")
### Make plot
ggplot(data = subset(user.data, date > jan1)) +
geom_point(aes(y=rank_newest,x=date, 
               color=point.label, alpha=point.label, 
               size=point.label)) + 
ggtitle("Distribution of Post Dates across Authors\nbetween Jan. and Oct.") +
ylab("Author Rank by Newest Post") + xlab("Date") +
geom_vline(aes(xintercept = as.numeric(july1)), alpha=0.3) + 
geom_vline(aes(xintercept = as.numeric(june1)), alpha=0.3) + 
scale_colour_manual(values = c("grey","darkblue", "red")) + 
scale_size_manual(values = c(0.1,0.1,0.3)) + 
scale_alpha_manual(values = c(0.1,0.1,1)) + 
theme_bw() + facet_wrap("si.group")
```
**Figure 1:** *Post histories for SI-Positive authors and SI-Negative authors are different.* 
The post histories for our authors are depicted above. Each y value represents a unique author id, sorted by the time of their most recent post. Each x value represents the date of a post. Hence, we can read each horizontal line on this plot as the post history of one author. Red points represent SI-positive posts. Blue points represent posts in the /r/depression subreddit. Grey posts represent posts elsewhere on Reddit.

There a number of important features in this plot. First, we note that those authors who posted more recently tended to post more often, which is expected. Second we see that many authors made multiple SI-positive posts. Third, we see that nearly 170 of our 475 authors had their last post in June, which was the period that we used to sample from. Fourth, of these 170 authors, many did not post frequently before dropping-off Reddit. However, a hand-full of them had posted frequently, and their drop-put is therefore suspicious.

@@

```{r, echo=FALSE, warning=FALSE}
load("Data/Posteriors/Attributable_10e-10.Rdata")
hist(post.excess, 
     main = "Posterior Distribution for\nExpected Reddit Drop-Outs per 100 SI-Positive Authors\n(Prior = 10e-4)",
     xlab="", ylab="")
ci = quantile(post.excess, probs = c(0.025, 0.975))
```

@@

We found that our conclusions were sensitive to our specified prior probability of drop-outs. The plot below shows posterior credible intervals for the expected number of excess dropouts among authors who penned SI-positive posts in June /r/depression.

@@

```{r, echo=FALSE, warning=FALSE}
      load("Data/Posteriors/Attributable_All.Rdata")
      load("Data/Posteriors/Priors_All.Rdata")
      ggplot() +
            geom_errorbar(aes(ymax = post_p[3,], ymin = post_p[1,], x=prior.p)) +
            geom_point(aes(y= post_p[2,], x=prior.p)) + 
            scale_x_log10() + 
            xlab("Prior Probability of Quitting Reddit") + 
            ylab("Excess Droup-Outs Associated with SI") +
            ggtitle("Posterior Credible Intervals for Total SI-Associated Drop-Outs") 
```
